### 2020-06-24 latex/vnhcs.tex
Continued writing simply actuated mech systems section, almost there.

Lagrangian mechanics uses the kinetic energy \(T(q,\dot{q})\) and potential
energy \(P(q)\) of the system to define the Lagrangian
(\ref{eqn:lagrangian-general})
\begin{equation}\label{eqn:lagrangian-general}
    \mathcal{L}(q,\dot{q}) = T(q,\dot{q}) - P(q)
\end{equation}
In actuated mechanical systems, the EOM are described by \(n\) second-order
ordinary differential equations (ODEs) obtained from the \textit{Euler-Lagrange
equations} with \textit{generalized input forces} \(\tau \in \R^m\) 
(\ref{eqn:el-eqns-general}). 
\begin{equation}\label{eqn:el-eqns-general}
    \diff{}{t}\left\{ \pdiff{\mathcal{L}}{\dot{q}_i} \right\}
    - \pdiff{\mathcal{L}}{q_i} = B_i(q) \tau
\end{equation}
Here, \(B_i: \mathcal{Q} \rightarrow \R^{1\times m}\)
and the matrix  \(B: \mathcal{Q} \rightarrow \R^{n \times m}\) given by
\[
    B(q) =\begin{bmatrix}
        - & B_1(q) & - \\
          & \vdots & \\
        - & B_n(q) & -
    \end{bmatrix}
\]
is the \textit{input matrix} for the system. If \(m < n\), we say the system is
\textit{underactuated} with degree of underactuation \(n - m\).

Many actuated mechanical systems have quadratic kinetic energies, so that the
Lagrangian can be written explicitly as
\begin{equation}\label{eqn:lagrangian}
    \mathcal{L}(q,\dot{q}) = \frac{1}{2} \dot{q}\tpose D(q) \dot{q} - P(q)
\end{equation}
where the \textit{inertia matrix} \(D: \mathcal{Q} \rightarrow \R^{n\times n}\) 
is a symmetric, positive definite matrix for all \(q \in \mathcal{Q}\) and the
potential function \(P : \mathcal{Q} \rightarrow \R\) is smooth. 
If this is the case, the Euler-Lagrange equations
(\ref{eqn:el-eqns-general}) reduce to (\ref{eqn:el-eqns}),
\begin{equation}\label{eqn:el-eqns}
    D(q)\ddot{q} + C(q,\dot{q})\dot{q} + \nabla P(q) = B(q)\tau
\end{equation}
where the \textit{Coriolis matrix} \(C(q,\dot{q})\) is of the form
\[
    [C]_{i,j} = \frac{1}{2}\sum\limits_{k = 1}^n 
    \left(\pdiff{[D]_{i,j}}{q_k}  +
     \pdiff{[D]_{i,k}}{q_j} -
     \pdiff{[D]_{k,j}}{q_i}\right)\dot{q}_k
\]

@@

Next we define the \textit{Hamiltonian} (\ref{eqn:hamiltonian-general}) 
as the Legendre transform of the Lagrangian \textbf{SOURCE}:
\begin{equation}\label{eqn:hamiltonian-general}
    \mathcal{H}(q,p) = p\tpose \dot{q}(q,p) - \mathcal{L}(q,\dot{q}(q,p))
\end{equation}
The EOM in this framework can be shown to be the \(2n\)
first-order equations
\begin{equation}\label{eqn:hamiltonian-eom-general}
    \begin{cases}
        \dot{q} = \pdiff{\mathcal{H}}{p} \\
        \dot{p} = -\pdiff{\mathcal{H}}{q} + B(q)\tau \\
    \end{cases}
\end{equation}
where \(B(q) \in \R^{n\times m}\) is the input matrix and 
\(\tau \in \R^m\) is the vector of generalized input forces.

If the Lagrangian is given as in (\ref{eqn:lagrangian}), the conjugate of
momenta is computed explicitly by
\[
    p = D(q)\dot{q}
\]
which gives the Hamiltonian (\ref{eqn:hamiltonian})
\begin{equation}\label{eqn:hamiltonian}
    \mathcal{H}(q,p) = \frac{1}{2} p\tpose D^{-1}(q) p + P(q)
\end{equation}
and the EOM from (\ref{eqn:hamiltonian-eom-general}) reduce to 
(\ref{eqn:hamiltonian-eom}).
\begin{equation}\label{eqn:hamiltonian-eom}
    \begin{cases}
        \dot{q} = D^{-1}(q)p \\
        \dot{p_i} = -\frac{1}{2} p\tpose \pdiff{D^{1}}{q_i}(q) p -
        \pdiff{P}{q_i}(q) + B_i(q) \tau \\
    \end{cases}
\end{equation}

\section{Simply Actuated Hamiltonian Systems}
% TODO: Describe the change of coordinates to get into q_u/q_a mode and show
% that the actuator directly affects pa but not pu. Use M for simply actuated
% inertia, D for normal coordinates
Given a mechanical system with phase coordinates \((q,p)\) and Hamiltonian
(\ref{eqn:hamiltonian}), it is not in general obvious how the input forces
\(\tau\) are affecting the conjugate of momenta since they are scaled by
\(B(q)\), which may be quite complicated. Even if \(B(q)\) is constant, multiple
forces may affect each \(p_i\). This section will demonstrate a useful change of
coordinates which results in a simpler input matrix.

@@

To perform our coordinate transformation, we define a transformation matrix
\[
    \mathbf{B} = 
    \begin{bmatrix}
        B^\perp \\
        B\tpose \\
    \end{bmatrix}
\]
Since both \(B^\perp\) and \(B\tpose\) are right-semi-orthogonal, 
\(\mathbf{B}\) is orthogonal. That is,
\(\mathbf{B}\mathbf{B}\tpose = \Id{n}\) which means
\(\mathbf{B}^{-1} = \mathbf{B}\tpose\).

\begin{defn}
    We call \(q_u = B^\perp q\) the \textbf{unactuated coordinates} and 
    \(q_a = B\tpose\) the \textbf{actuated coordinates}, which are obtained
    through the transformation
    \[
        \begin{bmatrix} q_u \\ q_a \end{bmatrix} = \mathbf{B}q
    \]
\end{defn}

>>>>

Lagrangian mechanics uses the kinetic energy \(T(q,\dot{q})\) and potential
energy \(P(q)\) of the system to define the Lagrangian
(\ref{eqn:lagrangian-general})
\begin{equation}\label{eqn:lagrangian-general}
    \mathcal{L}(q,\dot{q}) = T(q,\dot{q}) - P(q)
\end{equation}
In actuated mechanical systems, the EOM are described by \(n\) second-order
ordinary differential equations (ODEs) obtained from the \textit{Euler-Lagrange
equations} with \textit{generalized input forces} \(\tau \in \R^m\) 
(\ref{eqn:el-eqns-general}). 
\begin{equation}\label{eqn:el-eqns-general}
    \diff{}{t}\left\{ \pdiff{\mathcal{L}}{\dot{q}_i} \right\}
    - \pdiff{\mathcal{L}}{q_i} = B_i(q) \tau
\end{equation}
Here, \(B_i: \mathcal{Q} \rightarrow \R^{1\times m}\)
and the matrix  \(B: \mathcal{Q} \rightarrow \R^{n \times m}\) given by
%\[
%    B(q) =\begin{bmatrix}
%        - & B_1(q) & - \\
%          & \vdots & \\
%        - & B_n(q) & -
%    \end{bmatrix}
%\]
\[
    B(q) = [B_i(q)]
\]
is the \textit{input matrix} for the system. If \(m < n\), we say the system is
\textit{underactuated} with degree of underactuation \(n - m\).

Many actuated mechanical systems have quadratic kinetic energies, so that the
Lagrangian can be written explicitly as
\begin{equation}\label{eqn:lagrangian}
    \mathcal{L}(q,\dot{q}) = \frac{1}{2} \dot{q}\tpose D(q) \dot{q} - P(q)
\end{equation}
where the \textit{inertia matrix} \(D: \mathcal{Q} \rightarrow \R^{n\times n}\) 
is a symmetric, positive definite matrix for all \(q \in \mathcal{Q}\) and the
potential function \(P : \mathcal{Q} \rightarrow \R\) is smooth. 
If this is the case, the Euler-Lagrange equations
(\ref{eqn:el-eqns-general}) reduce to (\ref{eqn:el-eqns}),
\begin{equation}\label{eqn:el-eqns}
    D(q)\ddot{q} + C(q,\dot{q})\dot{q} + \nabla P(q) = B(q)\tau
\end{equation}
where the \textit{Coriolis matrix} \(C(q,\dot{q})\) is of the form
\[
    [C]_{i,j} = \frac{1}{2}\sum\limits_{k = 1}^n 
    \left(\pdiff{D_{i,j}}{q_k}  +
     \pdiff{D_{i,k}}{q_j} -
     \pdiff{D_{k,j}}{q_i}\right)\dot{q}_k
\]

@@

Next we define the \textit{Hamiltonian} (\ref{eqn:hamiltonian-general}) 
as the Legendre transform of the Lagrangian \textbf{SOURCE}:
\begin{equation}\label{eqn:hamiltonian-general}
    \mathcal{H}(q,p) = p\tpose \dot{q}(q,p) - \mathcal{L}(q,\dot{q}(q,p))
\end{equation}
The EOM in this framework can be shown to be the \(2n\)
first-order equations called \textit{Hamilton's equations}
\begin{equation}\label{eqn:hamiltons-eqns}
    \begin{cases}
        \dot{q} = \nabla_p\mathcal{H} \\
        \dot{p} = -\nabla_q\mathcal{H} + B(q)\tau \\
    \end{cases}
\end{equation}
where \(B(q) \in \R^{n\times m}\) is the input matrix and 
\(\tau \in \R^m\) is the vector of generalized input forces. A pair of
coordinates \((q,p)\) which satisfy Hamilton's equations are said to be 
\textit{canonical coordinates for} \(\mathcal{H}\).

If the Lagrangian is given as in (\ref{eqn:lagrangian}), the conjugate of
momenta is computed explicitly by
\[
    p = D(q)\dot{q}
\]
which gives the Hamiltonian (\ref{eqn:hamiltonian})
\begin{equation}\label{eqn:hamiltonian}
    \mathcal{H}(q,p) = \frac{1}{2} p\tpose D^{-1}(q) p + P(q)
\end{equation}
and the EOM from (\ref{eqn:hamiltons-eqns}) reduce to 
(\ref{eqn:hamiltonian-eom}).
\begin{equation}\label{eqn:hamiltonian-eom}
    \begin{cases}
        \dot{q} = D^{-1}(q)p \\
        \dot{p} = -\frac{1}{2} \left[p\tpose \pdiff{D^{-1}}{q_i}(q) p\right] 
        - \nabla_q P(q) + B(q) \tau \\
    \end{cases}
\end{equation}

\section{Simply Actuated Hamiltonian Systems}
% TODO: Describe the change of coordinates to get into q_u/q_a mode and show
% that the actuator directly affects pa but not pu. Use M for simply actuated
% inertia, D for normal coordinates
Given a mechanical system with Lagrangian (\ref{eqn:lagrangian}) yielding 
Hamiltonian (\ref{eqn:hamiltonian}) and canonical coordinates \((q,p)\), 
it is not in general obvious how the input forces
\(\tau\) are affecting the conjugate of momenta since they are scaled by
\(B(q)\) (which may be quite complicated). Even if \(B(q)\) is constant, multiple
forces may affect each \(p_i\). This section will demonstrate a useful change of
coordinates which results in a new Hamiltonian mechanical system with simpler
input matrix.

@@

To perform our coordinate transformation, we define a transformation matrix
\[
    \mathbf{B} = 
    \begin{bmatrix}
        B^\perp \\
        B\tpose \\
    \end{bmatrix}
\]
Since both \(B^\perp\) and \(B\tpose\) are right-semi-orthogonal and
\(\mathbf{B} \in \R^{n \times n}\) is full rank, \(\mathbf{B}\) is orthogonal. 
That is, \(\mathbf{B}\mathbf{B}\tpose = \Id{n}\) which means 
\(\mathbf{B}^{-1} = \mathbf{B}\tpose\).

\begin{defn}
    Define \(q_u = B^\perp q \in \R^{(n-k)}\) and 
    \(q_a = B\tpose q \in \R^k\). We say \(q_u\) are the \textit{unactuated
    coordinates} while \(q_a\) are the \textit{actuated coordinates} of the
    system. Their corresponding conjugate of momenta are labelled \(p_u\) and
    \(p_a\).
\end{defn}

In these new coordinates the input matrix will be modified such
that \(\tau_i\) only affects \(p_{a,i}\), and all \(p_u\) are unaffected by the
inputs.

To derive this result, observe that the coordinates 
\(\hat{q} = (q_u,q_a)\) can be written 
as the transformation (\ref{eqn:q-transformation}).
\begin{equation}\label{eqn:q-transformation}
    \hat{q} = \mathbf{B}q
\end{equation}
Since \(\mathbf{B}\) is
an orthogonal invertible matrix, we can equivalently write this as
\(q = \mathbf{B}\tpose \hat{q}\). Applying this change of coordinates to the
Lagrangian (\ref{eqn:lagrangian}) gives the new Lagrangian
\[
    \hat{\mathcal{L}}(\hat{q},\dot{\hat{q}}) = 
    \frac{1}{2}\dot{\hat{q}}\tpose \mathbf{B} 
    D(\mathbf{B}\tpose\hat{q}) 
    \mathbf{B}\tpose\dot{\hat{q}} - P(\mathbf{B}\hat{q})
\]
From here it is straightforward to show that the 
conjugate of momenta \(\hat{p} = (p_u,p_a)\) to \(\hat{q}\) is given by 
(\ref{eqn:p-transformation}).
\begin{equation}\label{eqn:p-transformation}
    \hat{p} = \pdiff{\hat{\mathcal{L}}}{\dot{\hat{q}}}
            = \mathbf{B}D(q)\dot{q} = \mathbf{B}p
\end{equation}

For ease of notation, we define
\(M(\hat{q}) := \mathbf{B}D\left(\mathbf{B}\tpose\hat{q}\right)\mathbf{B}\tpose\)
and 
\(V(\hat{q}) := P\left(\mathbf{B}\tpose \hat{q}\right)\)
Then the Hamiltonian (\ref{eqn:hamiltonian}) is given in
\((\hat{q},\hat{p})\) coordinates by
\begin{align*}
    \hat{\mathcal{H}}(\hat{q},\hat{p})
    &=
    \frac{1}{2}\hat{p}\tpose\mathbf{B}
    D^{-1}\left(\mathbf{B}\tpose\hat{q}\right)
    \mathbf{B}\tpose\hat{p}
    + P\left(\mathbf{B}\hat{q}\right) \\
    &= \hat{p}\tpose \Minv(\hat{q})\hat{p} + V(\hat{q})
\end{align*}

Now we will show \((\hat{q},\hat{p})\) are canonical coordinates
for \(\hat{\mathcal{H}}\). From the definition of \(\hat{q}\) in
(\ref{eqn:q-transformation}), we obtain
\begin{align*}
    \dot{\hat{q}} &= \mathbf{B}\dot{q} = \mathbf{B}D^{-1}(q)p \\
        &= \mathbf{B}D^{-1}(\mathbf{B}\tpose \hat{q})\mathbf{B}\tpose \hat{p} \\
        &= \Minv(\hat{q})\hat{p} \\
        &= \nabla_{\hat{p}}\hat{\mathcal{H}}
\end{align*}

The dynamics of \(\hat{p}\) are given by
\[
    \dot{\hat{p}}_i = \mathbf{B}\dot{p}
                    = -\frac{1}{2}\mathbf{B}
                    \left[p\tpose \pdiff{D^{-1}}{q_i}p\right]
                    -\mathbf{B} \nabla_q P(q) + \mathbf{B}B \tau
\]
where 
\[
    \mathbf{B} \nabla_q P(q)
    = \nabla_{\hat{q}}(\mathbf{B}\tpose \hat{q})\nabla_q P(q)
    = \nabla_{\hat{q}}q \nabla_q P(q)
    = \nabla_{\hat{q}}P(\mathbf{B}\tpose\hat{q}) 
    = \nabla_{\hat{q}} V(\hat{q}) 
\]

\textbf{LOOK INTO SYMPLECTIc jacobian method for canonical transformations to
show this is a canonical transformation}
